---
title: "reverse_engineer_#3"
author: "Victoria Stavish and Parker Leipzig"
date: "2022-11-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## project overview 

## Our project looks at homicide arrests in 50 of this nation's largest cities across the past decade. We used this data to replicate findings from a Washington Post investigation into arrest rates of homicicde cases in the largest U.S. cities. In order to achieve this, we first analyzed the Washington Post investigation and identified the most significant data findings within the story. Once we identified the most significant data findings, we created this notebook to recreate those significant findings. Here are some of the most significant findings we replicated:
1. There was a killing nearly every day between 2015-2017 in Baltimore
2. About one in every four homicides in Baltimore resulted in arrests 2015 - 2017
3. Baltimore's homicide arrest rate dropped 14 percentage points between 2015 and 2017

## data dictionary 

## The only dataset being used to replicate the Washington Post findings is titled 'homicide_data' in this notebook. This dataset has 52,179 entries and has 12 columns.
col1: uid is the unique identification number for each homicide case
col2: reported_date is the year, month and day the homicide was reported
col3: victim_last is the homicide victim's last name
col4: victim_first is the homicide victim's first name
col5: victim_race is the homicide victim's race
col6: victim_age is the homicide victim's age
col7: victim_sex is the homicide victim's gender
col8: city is the city the homicide occurred in
col9: state is the state the homicide occurred in 
col10: lat is the latitude the homicide occurred at 
col11: lon is the longitude the homicide occurred at
col12: disposition is the outcome of the case

# load libraries

```{r}

library(tidyverse)
library(janitor)
library(sf)

```

# load data

```{r}

homicide_data <- read_csv("https://raw.githubusercontent.com/washingtonpost/data-homicides/master/homicide-data.csv")


```

# Finding 1

# almost a killing each day for the past three years in Baltimore

# There are 1,095 days between 01/01/2015 and 12/31/2017. There were 1,002 reported homicides in Baltimore, MD during that same time period, proving that there was almost a killing each day for the past three years (imagining that we are writing this in 2018) in Baltimore, MD

# By Victoria 

```{r}
# find the amount of killings in 3 years 

baltimore_daily <- homicide_data %>%
  clean_names() %>%
  as.data.frame() %>%
  filter(city =="Baltimore") %>%
  select(uid, reported_date, city, state, disposition) %>%
  mutate(reported_date = lubridate::ymd(reported_date)) %>%
  mutate(year = lubridate::year(reported_date)) %>%
  filter(grepl("2015|2016|2017", year))

```

# Finding 2

# City police made an arrest in 41 percent of homicides in 2014 in Baltimore

# City police made an arrest in 40.76 percent of homicides in 2014 in Baltimore

# By Victoria 

```{r}
# what percent of Baltimore homicides resulted in arrests in 2014

bmore_2014_arrests <- homicide_data %>%
  clean_names() %>%
  as.data.frame() %>%
  filter(city =="Baltimore") %>%
  select(uid, reported_date, city, state, disposition) %>%
  mutate(reported_date = lubridate::ymd(reported_date)) %>%
  mutate(year = lubridate::year(reported_date)) %>%
  filter(grepl("2014", year)) %>%
  group_by(disposition) %>%
  count() %>%
  pivot_wider(names_from=disposition, values_from=n) %>%
  clean_names() %>%
  mutate(pct_unsolved = (closed_by_arrest / (closed_by_arrest + closed_without_arrest + open_no_arrest)))
  
```

# Finding 3

# last year (2017), the arrest rate was just 27 percent, a 14 percentage point drop in Baltimore

# City police made an arrest in 27.35 percent of homicides in 2017 in Baltimore

# By Victoria 

```{r}

# what percent of Baltimore homicides resulted in arrests in 2017

bmore_2017_arrests <- homicide_data %>%
  clean_names() %>%
  as.data.frame() %>%
  filter(city =="Baltimore") %>%
  select(uid, reported_date, city, state, disposition) %>%
  mutate(reported_date = lubridate::ymd(reported_date)) %>%
  mutate(year = lubridate::year(reported_date)) %>%
  filter(grepl("2017", year)) %>%
  group_by(disposition) %>%
  count() %>%
  pivot_wider(names_from=disposition, values_from=n) %>%
  clean_names() %>%
  mutate(pct_unsolved = (closed_by_arrest / (closed_by_arrest + closed_without_arrest + open_no_arrest)))

```

# Finding 4

# Of 50 of the nation’s largest cities, Baltimore is one of 34 where police now (2017) make homicide arrests less often than in 2014

# 27% of Baltimore homicides resulted in arrests in 2017 while 41% of Baltimore homicides resulted in arrests in 2014

# By Victoria

```{r}

# finding the 34 cities where police make less arrests in 2017 than in 2014

top_34_17 <- homicide_data %>%
  clean_names() %>%
  as.data.frame() %>%
  select(uid, reported_date, city, state, disposition) %>%
  mutate(reported_date = lubridate::ymd(reported_date)) %>%
  mutate(year = lubridate::year(reported_date)) %>%
  filter(grepl("2017", year)) %>%
  group_by(disposition, city, year) %>%
  count() %>%
  pivot_wider(names_from=disposition, values_from=n) %>%
  clean_names() %>%
  mutate(pct_unsolved17 = (closed_by_arrest / (closed_by_arrest + closed_without_arrest + open_no_arrest))) %>%
  drop_na() %>%
  arrange(desc(pct_unsolved17)) %>%
  head(top_34_17,n=34)


top_34_14 <- homicide_data %>%
  clean_names() %>%
  as.data.frame() %>%
  select(uid, reported_date, city, state, disposition) %>%
  mutate(reported_date = lubridate::ymd(reported_date)) %>%
  mutate(year = lubridate::year(reported_date)) %>%
  filter(grepl("2014", year)) %>%
  group_by(disposition, city, year) %>%
  count() %>%
  pivot_wider(names_from=disposition, values_from=n) %>%
  clean_names() %>%
  mutate(pct_unsolved14 = (closed_by_arrest / (closed_by_arrest + closed_without_arrest + open_no_arrest))) %>%
  drop_na() %>%
  arrange(desc(pct_unsolved14)) %>%
  head(top_34_14,n=34)

# joing the 2017 and 2014 dataframes and finding pct change

full_top_34 = merge(x=top_34_17,y=top_34_14,by="city",all=TRUE) %>%
  mutate(pct_diff = pct_unsolved17 - pct_unsolved14)

# baltimore arrest rate in 2017

bmore_2017_arrests <- homicide_data %>%
  clean_names() %>%
  as.data.frame() %>%
  filter(city =="Baltimore") %>%
  select(uid, reported_date, city, state, disposition) %>%
  mutate(reported_date = lubridate::ymd(reported_date)) %>%
  mutate(year = lubridate::year(reported_date)) %>%
  filter(grepl("2017", year)) %>%
  group_by(disposition) %>%
  count() %>%
  pivot_wider(names_from=disposition, values_from=n) %>%
  clean_names() %>%
  mutate(pct_unsolved = (closed_by_arrest / (closed_by_arrest + closed_without_arrest + open_no_arrest)))

# baltimore arrest rate in 2014

bmore_2014_arrests <- homicide_data %>%
  clean_names() %>%
  as.data.frame() %>%
  filter(city =="Baltimore") %>%
  select(uid, reported_date, city, state, disposition) %>%
  mutate(reported_date = lubridate::ymd(reported_date)) %>%
  mutate(year = lubridate::year(reported_date)) %>%
  filter(grepl("2014", year)) %>%
  group_by(disposition) %>%
  count() %>%
  pivot_wider(names_from=disposition, values_from=n) %>%
  clean_names() %>%
  mutate(pct_unsolved = (closed_by_arrest / (closed_by_arrest + closed_without_arrest + open_no_arrest)))

```

# Finding 5

# In Chicago, the homicide arrest rate has dropped 21 percentage points (2014 compared to 2017)

# in 2014, Chicago's homicide arrest rate was 29.90 percent. In 2017, Chicago's homicide arrest rate was 8.56 percent.

# 29.90 - 8.56 = 21.34 percentage point drop in homicide arrest rates

# By Victoria 

```{r}

# What was the homicide arrest rate in Chicago in 2014

chi_2014_arrests <- homicide_data %>%
  clean_names() %>%
  as.data.frame() %>%
  filter(city =="Chicago") %>%
  select(uid, reported_date, city, state, disposition) %>%
  mutate(reported_date = lubridate::ymd(reported_date)) %>%
  mutate(year = lubridate::year(reported_date)) %>%
  filter(grepl("2014", year)) %>%
  group_by(disposition) %>%
  count() %>%
  pivot_wider(names_from=disposition, values_from=n) %>%
  clean_names() %>%
  mutate(pct_unsolved = (closed_by_arrest / (closed_by_arrest + closed_without_arrest + open_no_arrest)))

# What was the homicide arrest rate in Chicago in 2017

chi_2017_arrests <- homicide_data %>%
  clean_names() %>%
  as.data.frame() %>%
  filter(city =="Chicago") %>%
  select(uid, reported_date, city, state, disposition) %>%
  mutate(reported_date = lubridate::ymd(reported_date)) %>%
  mutate(year = lubridate::year(reported_date)) %>%
  filter(grepl("2017", year)) %>%
  group_by(disposition) %>%
  count() %>%
  pivot_wider(names_from=disposition, values_from=n) %>%
  clean_names() %>%
  mutate(pct_unsolved = (closed_by_arrest / (closed_by_arrest + closed_without_arrest + open_no_arrest)))

```

# Finding 6

# In Boston, the homicide arrest rate has dropped 12 points 2014 - 2017

# In 2014 Boston's homicide arrest rate was 52.72 percent. In 2017 Boston's homicide arrest rate was 40.74 percent.

# 52.72 - 40.74 = 11.98 percentage point drop in Boston's homicide arrest rates

# BY Victoria 

```{r}

# What was the homicide arrest rate in Boston in 2017

bost_2017_arrests <- homicide_data %>%
  clean_names() %>%
  as.data.frame() %>%
  filter(city =="Boston") %>%
  select(uid, reported_date, city, state, disposition) %>%
  mutate(reported_date = lubridate::ymd(reported_date)) %>%
  mutate(year = lubridate::year(reported_date)) %>%
  filter(grepl("2017", year)) %>%
  group_by(disposition) %>%
  count() %>%
  pivot_wider(names_from=disposition, values_from=n) %>%
  clean_names() %>%
  mutate(pct_unsolved = (closed_by_arrest / (closed_by_arrest + open_no_arrest)))

# What was the homicide arrest rate in Boston in 2014

bost_2014_arrests <- homicide_data %>%
  clean_names() %>%
  as.data.frame() %>%
  filter(city =="Boston") %>%
  select(uid, reported_date, city, state, disposition) %>%
  mutate(reported_date = lubridate::ymd(reported_date)) %>%
  mutate(year = lubridate::year(reported_date)) %>%
  filter(grepl("2014", year)) %>%
  group_by(disposition) %>%
  count() %>%
  pivot_wider(names_from=disposition, values_from=n) %>%
  clean_names() %>%
  mutate(pct_unsolved = (closed_by_arrest / (closed_by_arrest + open_no_arrest)))

```

# Finding 6

# in St. Louis the homicide arrest rate has dropped 9 percentage points 2014 - 2017

# in 2014, St. Louis' homicide arrest rate was 40.63 percent. In 2017, St. Louis' homicide arrest rate was 31.68 percent.

# 40.63 - 31.68 = 8.95 percentage point drop in homicide arrest rates

# By Victoria 

```{r}

# What was the homicide arrest rate in St. Louis in 2014

louis_2014_arrests <- homicide_data %>%
  clean_names() %>%
  as.data.frame() %>%
  filter(city == "St. Louis") %>%
  select(uid, reported_date, city, state, disposition) %>%
  mutate(reported_date = lubridate::ymd(reported_date)) %>%
  mutate(year = lubridate::year(reported_date)) %>%
  filter(grepl("2014", year)) %>%
  group_by(disposition) %>%
  count() %>%
  pivot_wider(names_from=disposition, values_from=n) %>%
  clean_names() %>%
  mutate(pct_unsolved = (closed_by_arrest / (closed_by_arrest + closed_without_arrest + open_no_arrest)))

# What was the homicide arrest rate in St. Louis in 2017

louis_2017_arrests <- homicide_data %>%
  clean_names() %>%
  as.data.frame() %>%
  filter(city =="St. Louis") %>%
  select(uid, reported_date, city, state, disposition) %>%
  mutate(reported_date = lubridate::ymd(reported_date)) %>%
  mutate(year = lubridate::year(reported_date)) %>%
  filter(grepl("2017", year)) %>%
  group_by(disposition) %>%
  count() %>%
  pivot_wider(names_from=disposition, values_from=n) %>%
  clean_names() %>%
  mutate(pct_unsolved = (closed_by_arrest / (closed_by_arrest + closed_without_arrest + open_no_arrest)))

```

# Finding 7

# For most of the decade before 2015, Baltimore’s annual homicide arrest rate hovered at about 40 percent

# The arrest rate for the first half of the decade before 2015 hovered around 40.37 percent.

# By Parker
```{r}

# What was the homicide arrest rate in Baltimore 2010-2014

before_2015 <- homicide_data %>% 
  clean_names() %>% 
  as.data.frame() %>% 
  filter(city =="Baltimore") %>%
  select(uid, reported_date, city, state, disposition) %>%
  mutate(reported_date = lubridate::ymd(reported_date)) %>%
  mutate(year = lubridate::year(reported_date)) %>%
  filter(grepl("2010|2012|2013|2014", year)) %>%
  group_by(disposition, year) %>%
  count() %>%
  pivot_wider(names_from=disposition, values_from=n) %>%
  clean_names() %>%
  mutate(arrest_rate = (closed_by_arrest / (closed_by_arrest + closed_without_arrest + open_no_arrest)*100))

```

# Finding 8

# Since 2015, the arrest rate hasn’t topped 30 percent in any year in Baltimore

# Yes, all of these arrest rates in Baltimore 2015-2017 are below 30 percent.

# By Parker
```{r}

# What was the homicide arrest rate in Baltimore in 2015-2017

rate_2015 <- homicide_data %>% 
  clean_names() %>% 
  as.data.frame() %>% 
  filter(city =="Baltimore") %>%
  select(uid, reported_date, city, state, disposition) %>%
  mutate(reported_date = lubridate::ymd(reported_date)) %>%
  mutate(year = lubridate::year(reported_date)) %>%
  filter(grepl("2015", year)) %>%
  group_by(disposition) %>%
  count() %>%
  pivot_wider(names_from=disposition, values_from=n) %>%
  clean_names() %>%
  mutate(arrest_rate = (closed_by_arrest / (closed_by_arrest + closed_without_arrest + open_no_arrest)*100))

rate_2016 <- homicide_data %>% 
  clean_names() %>% 
  as.data.frame() %>% 
  filter(city =="Baltimore") %>%
  select(uid, reported_date, city, state, disposition) %>%
  mutate(reported_date = lubridate::ymd(reported_date)) %>%
  mutate(year = lubridate::year(reported_date)) %>%
  filter(grepl("2016", year)) %>%
  group_by(disposition) %>%
  count() %>%
  pivot_wider(names_from=disposition, values_from=n) %>%
  clean_names() %>%
  mutate(arrest_rate = (closed_by_arrest / (closed_by_arrest + closed_without_arrest + open_no_arrest)*100))

rate_2017 <- homicide_data %>% 
  clean_names() %>% 
  as.data.frame() %>% 
  filter(city =="Baltimore") %>%
  select(uid, reported_date, city, state, disposition) %>%
  mutate(reported_date = lubridate::ymd(reported_date)) %>%
  mutate(year = lubridate::year(reported_date)) %>%
  filter(grepl("2017", year)) %>%
  group_by(disposition) %>%
  count() %>%
  pivot_wider(names_from=disposition, values_from=n) %>%
  clean_names() %>%
  mutate(arrest_rate = (closed_by_arrest / (closed_by_arrest + closed_without_arrest + open_no_arrest)*100))

```

# Finding 9

# Of the 1,002 homicides between 2015 and the beginning of this year, just 252 — one out of every four — resulted in an arrest in Baltimore

# Just 25.15 percent of homicide cases 2015-2017 in Baltimore resulted in an arrest

# By Parker
```{r}

# How many homicides in Baltimore resulted in arrests between 2015-2017

balt_2015_2018 <- homicide_data %>% 
  clean_names() %>% 
  as.data.frame() %>% 
  filter(city =="Baltimore") %>%
  select(uid, reported_date, city, state, disposition) %>%
  mutate(reported_date = lubridate::ymd(reported_date)) %>%
  mutate(year = lubridate::year(reported_date)) %>%
  filter(grepl("2015|2016|2017", year)) %>%
  group_by(disposition) %>%
  count() %>%
  pivot_wider(names_from=disposition, values_from=n) %>%
  clean_names() %>%
  mutate(arrest_rate = (closed_by_arrest / (closed_by_arrest + closed_without_arrest + open_no_arrest)*100))
```

# Finding 10

# Baltimore’s decline (in homicide arrest rates) was sudden — plummeting 15 percentage points in 2015

# Between 2014 and 2015, Baltimore's arrest rate dropped from 40.76 percent to 25.44 percent.

# By Parker
```{r}

# What was the homicide arrest rate in Baltimore in 2014 vs 2015

balt_diff_2015 <- homicide_data %>% 
  clean_names() %>% 
  as.data.frame() %>% 
  filter(city =="Baltimore") %>%
  select(uid, reported_date, city, state, disposition) %>%
  mutate(reported_date = lubridate::ymd(reported_date)) %>%
  mutate(year = lubridate::year(reported_date)) %>%
  filter(grepl("2014|2015", year)) %>%
  group_by(disposition, year) %>%
  count() %>%
  pivot_wider(names_from=disposition, values_from=n) %>%
  clean_names() %>%
  mutate(arrest_rate = (closed_by_arrest / (closed_by_arrest + closed_without_arrest + open_no_arrest)*100))

```

