---
title: "data_project: public transportation"
author: "Victoria Stavish and Parker Leipzig"
date: "2022-11-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Project Summary 

We want to understand public transportation use across different Baltimore neighborhoods and why public transportation use fluctuates among these neighborhoods. To do this, we are using open source data sets from Baltimore Neighborhood Indicator Alliance that are updated as of 2020. 

# The specific datasets we are using are (add in data dictionaries for each):

* Median household Income
* Percent of Population that Uses Public Transportation to Get to Work
* Percent of Population that Drove Alone to Work
* Percent of Households with No Vehicles Available
* Percent of Residents - White/Caucasian (Non-Hispanic)
* Percent of Residents - Black/African American (Non-Hispanic)
* Percent of Residents - Hispanic
* Percent of Residents - Asian (Non-Hispanic)

# Using these datasets, we aim to answer these five questions:

1. Which 5 neighborhoods have the highest and lowest percentage of public transportation use? 

2. What are the income demographics of the neighborhoods with the highest and lowest transportation use?

3. What are the race demographics of the neighborhoods with the highest and lowest transportation use?

4. Do the neighborhoods that have the highest rates of people driving alone to work also have the lowest rates of people who use public transportation to get to work?

5. Do the neighborhoods that have lowest rates of people driving alone to work also have the highest rates of people who use public transportation to get to work?

# load libraries

```{r}

library(tidyverse)
library(dplyr)
library(janitor)
library(lubridate)
library(readxl)
library(ggthemes)
library(sf)
library(leaflet)

```

# Load data

```{r}

public_transportation <- read_excel("public_transportation/public_transportation.xlsx")

drive_to_work <- read_excel("public_transportation/drive_to_work.xlsx")

no_car <- read_excel("public_transportation/no_car_access.xlsx")

median_income <- read_excel("public_transportation/median_income.xlsx")

pop_white <- read_excel("public_transportation/population_white.xlsx")

pop_black <- read_excel("population_black/population_black.xlsx")

pop_hisp <- read_excel("population_black/population_hispanic.xlsx")

pop_asian <- read_excel("population_black/population_asian.xlsx")

```

# Which 5 neighborhoods have the highest and lowest percentage of public transportation use and what are the income demographics of these neighborhoods?

# Summary 

#The five Baltimore neighborhoods that have the highest percentage of residents using public transportation are, in order from greatest percentage to lowest percentage, are Sandtown-Winchester/Harlem Park, Southwest Baltimore, Greater Rosemont, Poppleton/The Terraces/Hollins Market and Clifton-Berea. Even though these five neighborhoods have the highest percentage of public transportation use, the highest percentage is in Sandtown-Winchester/Harlem Park at only 38.5%.

#The five Baltimore neighborhoods that have the lowest percentage of residents using public transportation are, in order from lowest percentage to greatest percentage, are Canton, South Baltimore, Greater Roland Park/Poplar Hill, Fells Point and Cross-Country/Cheswolde. The lowest percentage of people using public transportation is in Canton with just 3.8%.

# Data dictionary
* community: Neighborhoods in Baltimore City (chr)
* public_transport: Percent of people in the neighborhood who regularly use public transportation to get to work (dbl)
* median_income: the median income of each neighborhood (chr)

# By Victoria

```{r}
# 5 Baltimore neighborhoods that use public transportation the most to get to work 

max_public_transportation <- public_transportation %>%
  clean_names() %>%
  rename(public_transport = x2020_data) %>%
  slice_max(public_transport, n=5)

# 5 Baltimore neighborhoods that use public transportation the least to get to work 

min_public_transportation <- public_transportation %>%
  clean_names() %>%
  rename(public_transport = x2020_data) %>%
  slice_min(public_transport, n=5)

# Output dataframes

max_public_transportation
min_public_transportation

# inner join neighborhoods that use public transportation the most with income data
median_income <- median_income %>%
  clean_names()

max_transit_income <- max_public_transportation %>%
  clean_names() %>%
  inner_join(median_income, by="community") %>%
  rename(median_income = x2020_data)

# inner join neighborhoods that use public transportation the least with income data

min_transit_income <- min_public_transportation %>%
  clean_names() %>%
  inner_join(median_income, by="community") %>%
  rename(median_income = x2020_data)

# output dataframes

max_transit_income
min_transit_income

```

# What are the race demographics of the neighborhoods with the highest and lowest transportation use?

# Summary 

#Neighborhoods that have the highest percentage of public transportation use are majority Black while the neighborhoods that have the lowest percentage of public transportation use are majority White. Additionally, the median incomes of the majority Black and high public transportation use incomes are significantly lower than the median incomes of the majority White and low public transportation use neighborhoods.

#Most of the low public transportation use White neighborhoods make near or over a median of $100,000 while none of the high public transportation and majority Black community make more than a median of $40,000

# Data dictionary
* community: Neighborhoods in Baltimore City (chr)
* public_transport: Percent of people in the neighborhood who regularly use public transportation to get to work (dbl)
* median_income: the median income of each neighborhood (chr)
* asian_pop: the percent of the population that is Asian in a given neighborhood (dbl)
* black_pop: the percent of the population that is Black in a given neighborhood (dbl)
* white_pop: the percent of the population that is White in a given neighborhood (dbl)
* hisp_pop: the percent of the population that is Hispanic in a given neighborhood (dbl)


# By Victoria

```{r}

# clean the names of all race dataframes so they join easily

pop_asian <- pop_asian %>%
  clean_names() 
 
pop_white <- pop_white %>%
  clean_names()

pop_black <- pop_black %>%
  clean_names() 

pop_hisp <- pop_hisp %>%
  clean_names() 
  
# inner join race dataframes with the top public transport use and income dataframe

max_transit_race <- max_transit_income %>%
  clean_names() %>%
  inner_join(pop_asian, by="community") %>%
  inner_join(pop_black, by="community") %>%
  inner_join(pop_white, by="community") %>%
  inner_join(pop_hisp, by="community")
names(max_transit_race)[names(max_transit_race) == "x2020_data.x"] <- "asian_pop"
names(max_transit_race)[names(max_transit_race) == "x2020_data.y"] <- "black_pop"
names(max_transit_race)[names(max_transit_race) == "x2020_data.x.x"] <- "white_pop"
names(max_transit_race)[names(max_transit_race) == "x2020_data.y.y"] <- "hisp_pop"


# inner join race dataframes with the minimum public transport use and income dataframe

min_transit_race <- min_transit_income %>%
  clean_names() %>%
  inner_join(pop_asian, by="community") %>%
  inner_join(pop_black, by="community") %>%
  inner_join(pop_white, by="community") %>%
  inner_join(pop_hisp, by="community")
names(min_transit_race)[names(min_transit_race) == "x2020_data.x"] <- "asian_pop"
names(min_transit_race)[names(min_transit_race) == "x2020_data.y"] <- "black_pop"
names(min_transit_race)[names(min_transit_race) == "x2020_data.x.x"] <- "white_pop"
names(min_transit_race)[names(min_transit_race) == "x2020_data.y.y"] <- "hisp_pop"

# Output dataframes 

max_transit_race
min_transit_race

```

# Do the neighborhoods that have the highest rates of car ownership also have the lowest rates of people who use public transportation to get to work?

# Summary 

In general, the Baltimore City neighborhoods that have the lowest percentage of people who use public transportation to get to work have some of the highest rates of car ownership. Usually between 3% - 11% of people do not own cars. 

# Data Dictionary
* community: Neighborhoods in Baltimore City (chr)
* public_transport: Percent of people in the neighborhood who regularly use public transportation to get to work (dbl)
* pct_wo_car: Percent of people in the neighborhood who do not have regular access/own a car


#By Parker

```{r}
drive_to_work <- drive_to_work %>%
  clean_names()

low_public_high_drive <- min_public_transportation %>% 
  clean_names() %>% 
  inner_join(drive_to_work, by="community") %>%
  rename(pct_wo_car = x2020_data)
  
low_public_high_drive

```

# Do the neighborhoods that have the highest rates of people who use public transportation to get to work also have the lowest rates of car ownership?

# Summary 

# In general, the Baltimore City neighborhoods that have the highest percentage of people who use public transportation to get to work also some of the lowest rates oof car ownership, usually between 40% - 50% of people do not own cars. 

# Data Dictionary
* community: Neighborhoods in Baltimore City (chr)
* public_transport: Percent of people in the neighborhood who regularly use public transportation to get to work (dbl)
* pct_wo_car: Percent of people in the neighborhood who do not have regular access/own a car

#By Parker
```{r}
no_car <- no_car %>%
  clean_names()

high_public_low_drive <- max_public_transportation %>% 
  clean_names() %>% 
  inner_join(no_car, by="community") %>%
  rename(pct_wo_car = x2020_data)

high_public_low_drive
```


# 250 words on newsworthy findings 

The communities with the highest use of public transportation are majority Black and the communities with the lowest use of public transport are majority white. For example, of the five neighborhoods with the highest percentage of public transportation use, two are over 90% Black, one is over 80% Black and two are over 70% Black. Of the five neighborhoods with the lowest percentage of public transportation use, one is over 80% white, two are over 70% white and two are over 60% white.  It is also notable that the neighborhoods with high percentage of public transportation use (which are also majority Black) have a median salary range of about $23,000 - $39,000 while the neighborhoods with low percentage of public transportation use (which are also majority white) have a median salary range of about $63,000 - $134,000. However, $63,000 is an outlier, and without that one neighborhood $97,000 - $134,000.


#WELLS COMMENT: Now, let's produce a map and a few charts to illustrate these findings.

#Map
```{r}
baltimore_shape <- st_read("/Users/parkerleipzig/Desktop/GitHub/reverse_engineer_project/csa_2010_boundaries (1)/CSA_NSA_Tracts.shp")

geo_public_transportation <- baltimore_shape %>% 
  inner_join(public_transportation, by="Community") %>% 
  clean_names() %>% 
  st_as_sf() %>% 
  rename(public_transportation_pct = x2020_data)

ggplot() +
  geom_sf(data=geo_public_transportation, aes(fill=public_transportation_pct)) +
  scale_fill_gradient(low="grey",high="blue") +
  labs(title = "Pct of People that Use Public Transportation in Balt Neighborhoods",
    subtitle ="Citywide average: 17.4 percent",
    caption = "BNIA Data 2020, Graphic by Parker Leipzig and Victoria Stavish",
      )

```
