---
title: "data_project: public transportation"
author: "Victoria Stavish and Parker Leipzig"
date: "2022-11-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Project Summary 

We want to understand public transportation use across different Baltimore neighborhoods and why public transportation use fluctuates among these neighborhoods. To do this, we are using open source data sets from Baltimore Neighborhood Indicator Alliance that are updated as of 2020. 

# The specific datasets we are using are (add in data dictionaries for each):

* Median household Income
* Percent of Population that Uses Public Transportation to Get to Work
* Percent of Population that Drove Alone to Work
* Percent of Households with No Vehicles Available
* Percent of Residents - White/Caucasian (Non-Hispanic)
* Percent of Residents - Black/African American (Non-Hispanic)
* Percent of Residents - Hispanic
* Percent of Residents - Asian (Non-Hispanic)

# Using these datasets, we aim to answer these five questions:

1. Which 5 neighborhoods have the highest and lowest percentage of public transportation use? 

2. What are the income demographics of the neighborhoods with the highest and lowest transportation use?

3. What are the race demographics of the neighborhoods with the highest and lowest transportation use?

4. Do the neighborhoods that have the highest rates of people driving alone to work also have the lowest rates of people who use public transportation to get to work?

5. Do the neighborhoods that have lowest rates of people driving alone to work also have the highest rates of people who use public transportation to get to work?

# load libraries

```{r}

library(tidyverse)
library(janitor)
library(lubridate)
library(readxl)
library(ggthemes)
library(sf)
library(leaflet)

```

# Load data

```{r}

public_transportation <- read_excel("public_transportation/public_transportation.xlsx")

drive_to_work <- read_excel("public_transportation/drive_to_work.xlsx")

no_car <- read_excel("public_transportation/no_car_access.xlsx")

median_income <- read_excel("public_transportation/median_income.xlsx")

pop_white <- read_excel("public_transportation/population_white.xlsx")

pop_black <- read_excel("population_black/population_black.xlsx")

pop_hisp <- read_excel("population_black/population_hispanic.xlsx")

pop_asian <- read_excel("population_black/population_asian.xlsx")

```

# Which 5 neighborhoods have the highest and lowest percentage of public transportation use? 

# By Victoria

```{r}
# 5 Baltimore neighborhoods that use public transportation the most to get to work 

max_public_transportation <- public_transportation %>%
  as.data.frame() %>%
  clean_names() %>%
  arrange(desc(x2020_data)) %>%
  slice_max(x2020_data, n=5)

# 5 Baltimore neighborhoods that use public transportation the least to get to work 

min_public_transportation <- public_transportation %>%
  as.data.frame() %>%
  clean_names() %>%
  slice_min(x2020_data, n=5)

# Output dataframes

max_public_transportation
min_public_transportation

```

#wells comment: write a sentence summarizing the key findings of these tables and why that is significant. i.e. Sandtown-Winchester/Harlem Park has the highest use of public transportation has this West Baltimore neighborhood historically has been among the poorest in the city. Canton has a lot of boats and Range Rovers. Enough said.


# What are the income demographics of the neighborhoods with the highest and lowest transportation use?

# By Victoria 

#Wells comment: This table below is good but it sort of renders the previous chunk at Line 82 obsolete. I would delete it and combine the questions here.
```{r}
# inner join neighborhoods that use public transportation the most with income data
median_income <- median_income %>%
  clean_names()

max_transit_income <- max_public_transportation %>%
  clean_names() %>%
  inner_join(median_income, by="community")
names(max_transit_income)[names(max_transit_income) == "x2020_data.x"] <- "high_public_transport"
names(max_transit_income)[names(max_transit_income) == "x2020_data.y"] <- "2020_income"

# inner join neighborhoods that use public transportation the least with income data

min_transit_income <- min_public_transportation %>%
  clean_names() %>%
  inner_join(median_income, by="community")
names(min_transit_income)[names(min_transit_income) == "x2020_data.x"] <- "low_public_transport"
names(min_transit_income)[names(min_transit_income) == "x2020_data.y"] <- "2020_income"

# output dataframes

max_transit_income
min_transit_income

```



# What are the race demographics of the neighborhoods with the highest and lowest transportation use?

# By Victoria (need to rename columns for clarity)

#Wells comment: This table is great but provide some analysis here about what we are looking at. Also, there are two "low_public_transport" columns so figure out which one to keep

```{r}

# clean the names of all race dataframes so they join easily

pop_asian <- pop_asian %>%
  clean_names() 
 
pop_white <- pop_white %>%
  clean_names()

pop_black <- pop_black %>%
  clean_names() 

pop_hisp <- pop_hisp %>%
  clean_names() 
  
# inner join race dataframes with the top public transport use and income dataframe

max_transit_race <- max_transit_income %>%
  clean_names() %>%
  inner_join(pop_asian, by="community") %>%
  inner_join(pop_black, by="community") %>%
  inner_join(pop_white, by="community") %>%
  inner_join(pop_hisp, by="community")
names(max_transit_race)[names(max_transit_race) == "x2020_data.x"] <- "high_public_transport"
names(max_transit_race)[names(max_transit_race) == "x2020_data.x"] <- "asian_pop"
names(max_transit_race)[names(max_transit_race) == "x2020_data.y"] <- "black_pop"
names(max_transit_race)[names(max_transit_race) == "x2020_data.x.x"] <- "white_pop"
names(max_transit_race)[names(max_transit_race) == "x2020_data.y.y"] <- "hisp_pop"


# inner join race dataframes with the minimum public transport use and income dataframe

min_transit_race <- min_transit_income %>%
  clean_names() %>%
  inner_join(pop_asian, by="community") %>%
  inner_join(pop_black, by="community") %>%
  inner_join(pop_white, by="community") %>%
  inner_join(pop_hisp, by="community")
names(min_transit_race)[names(min_transit_race) == "x2020_data.x"] <- "low_public_transport"
names(min_transit_race)[names(min_transit_race) == "x2020_data.x"] <- "asian_pop"
names(min_transit_race)[names(min_transit_race) == "x2020_data.y"] <- "black_pop"
names(min_transit_race)[names(min_transit_race) == "x2020_data.x.x"] <- "white_pop"
names(min_transit_race)[names(min_transit_race) == "x2020_data.y.y"] <- "hisp_pop"

# Output dataframes 

max_transit_race
min_transit_race

```

# Do the neighborhoods that have the highest rates of people driving alone to work also have the lowest rates of people who use public transportation to get to work?
#By Parker

#Wells comment: I don't understand what the column "no car" is supposed to measure. Canton is 79.3% no car? It seems like you should use the column names from the chunk at Line 222 which made more sense: "drives_to_work"

```{r}
drive_to_work <- drive_to_work %>%
  clean_names()

low_public_high_drive <- min_public_transportation %>% 
  clean_names() %>% 
  inner_join(drive_to_work, by="community") 

names(low_public_high_drive)[names(low_public_high_drive) == "x2020_data.x"] <- "low_public_transport"
names(low_public_high_drive)[names(low_public_high_drive) == "x2020_data.y"] <- "drive_to_work"
  

low_public_high_drive

```

# Do the neighborhoods that have lowest rates of people driving alone to work also have the highest rates of people who use public transportation to get to work?
#By Parker
```{r}
no_car <- no_car %>%
  clean_names()

high_public_low_drive <- max_public_transportation %>% 
  clean_names() %>% 
  inner_join(no_car, by="community") 

names(high_public_low_drive)[names(high_public_low_drive) == "x2020_data.x"] <- "high_public_transport"
names(high_public_low_drive)[names(high_public_low_drive) == "x2020_data.y"] <- "no_car"

high_public_low_drive
```


# 250 words on newsworthy findings 
The communities with the highest use of public transportation are majority Black and the communities with the lowest use of public transport are majority white. For example, of the five neighborhoods with the highest percentage of public transportation use, two are over 90% Black, one is over 80% Black and two are over 70% Black. Of the five #SPELLING# neighorhoods with the lowest percentage of public transportation use, one is over 80% white, two are over 70% white and two are over 60% white. #WELLS COMMENT: DO THESE NEIGHBORHOODS HAVE SUBWAY STOPS OR EASY ACCESS TO PUBLIC TRANSPORTATION INFRASTRUCTURE?## It is also notable that the neighborhoods with high percentage of public transportation use (which are also majority Black) have a median salary range of about #NEED DOLLAR SIGN!# 23,000 - 39,000 while the neighborhoods with low percentage of public transportation use (which are also majority white) have a median salary range of about #NEED DOLLAR SIGN!# 63,000 - 134,000. However, #NEED DOLLAR SIGN!# 63,000 is an outlier, and without that one neighborhood, the salary range is #NEED DOLLAR SIGN!# 97,000 - 134,000.

It's also notable that there is only one Baltimore neighborhood in which less than 40% of people live in a household with no vehicle access. Lot's of people in Baltimore, dpesite the specific neighborhood they live in, do not have cars. 


#WELLS COMMENT: Now, let's produce a map and a few charts to illustrate these findings.
